---
title: "0_Exploration_and_cleaning"
author: "Mark_Cheng"
date: "2023-12-15"
output: html_document
---

Key Goals: 
1. Remove preprints
2. Remove papers where cancer is a comorbidity, i.e. cancer vs non-cancer is not the main focus
3. Remove papers where cancer diagnosis recency is ambiguous (>1 year)
4. clean up the study IDs (last names)
5. remove newlines

# read in and Remove preprints
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, tidyr, stringr)

data_directory = "../data/"
data_directory_files = paste0(data_directory,list.files(data_directory))
# find the files - make sure they are in the data folder, and have these substrings
#DE_csv filename is edited when downloaded from covidence to ensure unique DE.csv suffix
DE_csv = data_directory_files[str_detect(data_directory_files,"DE.csv")]
included_list_csv = data_directory_files[str_detect(data_directory_files,"included_csv")]


# Read DE.csv file with preserving spaces and special characters in colnames
df_with_unknown_dx_with_preprints <- read.csv(DE_csv,header = TRUE, check.names = FALSE)
#remove_preprints
df_with_unknown_dx <- subset(df_with_unknown_dx_with_preprints,`Peer-reviewed or Preprint`!="Preprint")

df_with_unknown_dx
```

# Cleaning
```{r}
#inlcuded_df is all the screened and included papers, which includes the metadata status, most important being the tags
included_df <- read.csv(included_list_csv)
included_df

#removes cancer as comorbidity
included_df_true_c_vs_ns <- included_df[grepl("true cancer vs non-cancer",included_df$Tags),]

#which papers have ambiguous cancer diagnosis timing - AND NEEDS RE-INTERPRETATION?
ambiguous_cancer_substrings <- c("cancer diagnosis time unknown","subsets in cancer diagnosis recency")
ambiguous_cancer_papers <- included_df_true_c_vs_ns[grepl(paste(ambiguous_cancer_substrings, collapse = "|"), included_df_true_c_vs_ns$Tags), ] %>%
  filter(!str_detect(Tags,"univariate only"))
write.csv(ambiguous_cancer_papers,"../output/testing_different_sets/Ambiguous_cancer_papers_set.csv",row.names = FALSE)

#remove papers where cancer diagnosis is ambiguous, leaving only active cancers
included_df_true_c_vs_ns_active <- included_df_true_c_vs_ns %>% 
  filter(!str_detect(Tags,"cancer diagnosis time unknown \\(do not proceed"))

included_df_true_c_vs_ns_active

# use the above to remove DE_df with unknown diagnosis time
df <- df_with_unknown_dx[paste0("#",df_with_unknown_dx$`Covidence #`) %in% included_df_true_c_vs_ns_active$Covidence..,]
df
```

```{r}
# correct study ID
#correct Covidence ID inherited Study ID with extracted last name
df_rename <- df%>%
  mutate(first_author_last_name = str_split(`First Author (Last name and Initials)`,",| ",simplify=TRUE)[,1]) %>%
  select(`Study ID`,`Covidence #`,`First Author (Last name and Initials)`,first_author_last_name)

df_rename <- df_rename %>%
  mutate(year =str_extract(`Study ID`,"\\d{4}$")) %>%
  mutate(`Study ID` = ifelse(str_detect(`Study ID`, "^[A-Z] \\d{4}$"), #if follows the format of [initial] [year] that covidence allocates when can't detect last name
                                     paste0(first_author_last_name," ",year),
                                     `Study ID`))
## reassign corrected Study ID onto df
df$`Study ID` <- df_rename$`Study ID`

# remove newlines
df <- df %>%
  mutate_all(~gsub("\n", " ", .))

df
```


```{r}
#sanity check - the tags of papers in these included papers
included_df_true_c_vs_ns_active[paste0("#",df_with_unknown_dx$`Covidence #`) %in%
                                  included_df_true_c_vs_ns_active$Covidence..,]$Tags
# what papers are included by tag but does not have a consensus DE?
included_df_true_c_vs_ns_active[!included_df_true_c_vs_ns_active$Covidence.. %in%
                                  paste0("#",df$`Covidence #`),]$Covidence..
# sanity check - not included in DE but included by Tag should be 0
df[!paste0("#",df$`Covidence #`) %in% included_df_true_c_vs_ns_active$Covidence..,]
```

The papers included by tag but does not have a consensus DE are:
72854,
124497
are currently still in progress.
128722 is a preprint 2020
131660 is a peer-reviewed article, accidentally marked as preprint
135365	is a preprint 2022
136941	is a preprint 2020
137477	is a preprint 2020

## which papers are multivariate analysis only?
```{r}
# only include papers with multivariate analysis
df$`Univariate or multivariate analysis?` %>% table() %>% data.frame()


df_total <- df %>% group_by(`Univariate or multivariate analysis?`) %>% summarise(sum(as.numeric(`Total study size (number of patients)`)),sum(as.numeric(`Cancer Total number`)),sum(as.numeric(`Control (non-cancer) Total number`)))
df_total

write.csv(df_total,"../temporary_testing/total_uni_multi.csv") 
```

```{r}

df_count_outliers <- df[df$`Univariate or multivariate analysis?` %in% c("univariate","univariate; Other: crude","univariate; multivariate"),] %>%
  select(`Covidence #`,"Univariate or multivariate analysis?","Study ID","Total study size (number of patients)","Cancer Total number","Control (non-cancer) Total number")

df_count_outliers

df_count_outliers
write.csv(df_count_outliers,"../temporary_testing/total_uni_multi_NAs.csv")

## only pick up papers with multivariate
df <- df %>% filter(str_detect(`Univariate or multivariate analysis?`,"multivariate"))
```


# brief clean
```{r}
#replace p value with p-value so it doesn't get mixed with value[outcome]
colnames(df) <- gsub("p value","p-value",colnames(df))
#unify 95% CI and 95% confidence interval
colnames(df) <- gsub("95% confidence interval","95% CI",colnames(df))
#unify Measure of Association (OR/RR etc.)
colnames(df) <- gsub("Measure of Association \\(OR\\/RR etc.\\)","Measure of Association",colnames(df))
colnames(df) <- gsub("Measure of Association \\(OR/RR etc\\)","Measure of Association",colnames(df),ignore.case = TRUE)
colnames(df) <- gsub("Measure of Association \\(OR/RR\\)","Measure of Association",colnames(df),ignore.case = TRUE)
colnames(df) <- gsub("measurement of association \\(OR/RR\\)","Measure of Association",colnames(df),ignore.case = TRUE)

colnames(df) <- gsub("Measure of Association \\(OR/RR etc\\)","Measure of Association",colnames(df))
colnames(df) <- gsub("Measure of Association \\(OR/RR\\)","Measure of Association",colnames(df))
#unify event total
colnames(df) <- gsub("Cancer: Event/Total","cancer: Event/Total",colnames(df))
#unify Value and value
colnames(df) <- gsub("Value","value",colnames(df))
#unify { and [
colnames(df) <- gsub("\\{","[",colnames(df))

# Make column names unique without modifying spaces and special characters
df <- df %>%
  setNames(make.unique(names(.)))
df
```

```{r}
#exclude papers tagged explicitly with "univariate only"
univariate_only_list = included_df_true_c_vs_ns_active %>% 
  filter(str_detect(Tags,"univariate only")) %>%
  select(Covidence..) %>%
  mutate(Covidence.. = sub("^#","",Covidence..)) %>%
  as.list()
print(glue::glue("{univariate_only_list} papers with univariate only tag"))
#removal step
df <- df[!df$`Covidence #` %in% univariate_only_list$Covidence..,]
df
```



```{r}
#export
output_string <- sub("\\.csv$","_knownDx_peerReviewed.csv",DE_csv)

write.csv(df,file =output_string,row.names = FALSE)
```



