---
title: "2_reshaping_ORsRRs"
author: "Mark_Cheng"
date: "2023-12-19"
output: html_document
---

```{r}
# load libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, tidyr, stringr,readxl)

data_directory = "../data/"
data_directory_files = paste0(data_directory,list.files(data_directory))

DE_csv = data_directory_files[str_detect(data_directory_files,
                                         "_DE_knownDx_peerReviewed.csv")]
# initiate cleaned df
df <- read.csv(file = DE_csv, header = TRUE, check.names = FALSE)
df

## rearrange df to chronological order
df_location_date <- read.csv("../output/Timeline_and_Lineage_Imputation/Studies_split_Timeperiods_ordered_by_time_periods.csv",check.names = FALSE)
study_order <- df_location_date %>% distinct(`Covidence #`) %>% pull()

df <- df %>% arrange(match(`Covidence #`,study_order))
df
```


#Pseudocode
Observations for columns:
1. A-D Covidence automatic - Covidence ID, Study ID (Last Name + publication year), title
2. E-AA: General study information, definitions, study design, recruitment strategy
3. AB-AN: Baseline characteristics Table
4. AP-EH: Comorbidities Table
5. EI-EL: Definitions of outcomes of severity, ICU admission/critical, mortality, severe disease
6. EM: if univariate and/or multivariate is used, EN: What is included in multivariable analysis adjustments
7. Main Tables, Primary cancer site, metastasis status, treatment type, COVID variants, COVID vaccination status
8. each tables have the same repeating 27 columns for each row (Total + 6x columns for 4 outcomes + 2 comments).
Except for therapy type and COVID variant tables which has additional col2: PERCENTAGE

  col1:Total Number of Patients in that stratified group
  Mortality x6
  - Measure of Association (OR/RR etc.) [mortality]
  - value [mortality]
  - 95% confidence interval [mortality]
  - p value [mortality]
  - Cancer: Event/Total [mortality]
  - Non-cancer: Event/Total [mortality]
  ICU x6
  - Measure of Association (OR/RR etc.) [ICU]
  - value [ICU]
  - 95% confidence interval [ICU]
  - p value [ICU]
  - Cancer: Event/Total [ICU]
  - Non-cancer: Event/Total [ICU]
  severe disease x6
  - Measure of Association (OR/RR etc.) [severe disease]
  - value [severe disease]
  - 95% confidence interval [severe disease]
  - p value [severe disease]
  - Cancer: Event/Total [severe disease]
  - Non-cancer: Event/Total [severe disease]
  
  col20:Comment
  hospitalisation x6
  - Measure of Association (OR/RR etc.) [hospitalisation]
  - value [hospitalisation]	95% confidence interval [hospitalisation]
  - p value [hospitalisation]
  - Cancer: Event/Total [hospitalisation]
  - Non-cancer: Event/Total [hospitalisation]
  
  col27:Comment

# Only pick out papers with multivariate analysis - stop for now
```{r}
df$`Univariate or multivariate analysis?`
# df <- df[grepl(pattern = "multivariate",df$`Univariate or multivariate analysis?`),]
# df <- subset(df,`Covidence #`!=128789) #128789 does not actually provided mOR (crude OR/CI/p-values calculated as univariate/multivariate analysis not provided for outcomes of interest)
```
# Finding the tables
## Table of baseline characteristics
```{r}
baseline_start_col = which(colnames(df)=="Baseline characteristics:")
baseline_end_col = which(colnames(df)=="Which COVID-19 variant does the study identify?")
df_baseline_characteristics <- df%>%
  select(c(1, #keep Covidence ID
           ((baseline_start_col+1):(baseline_end_col-1))
           )
         )# gets the baseline characteristics table
df_baseline_characteristics

write.csv(df_baseline_characteristics,"../output/processed_tables/baseline_characteristics.csv",row.names = FALSE)
```

```{r how many cancer and non-cancer patients are there in total?}
df_baseline_characteristics$`Cancer Total number`

df_baseline_characteristics$`Control (non-cancer) Total number`
```
# find comorbidity table

```{r}
#remove MoA, OR/RR value, 95% confidence interval, and p value of comorbidities
comorb_start_col = which(colnames(df)=="Co-morbidities:") # column name before the comorbidity table
comorb_end_col = which(colnames(df)=="Definition of severity") #column name after the comorbidity table
df_comorbidity <- df %>%
  select(c(1,2,"Cancer Total number","Control (non-cancer) Total number",((comorb_start_col+1):(comorb_end_col-1)))) # de-select the comorbidity table

df_comorbidity

#for Abe
write.csv(df_comorbidity,"../output/abe_tables_graphs/ComorbidityTable_forAbe.csv")
```

# Result Tables
## Common Functions
### 1. Comment merging function
Function built to merge all comments to last instance in table, and remove all but last instance.
```{r collapse comments}
collapse_comment_to_last_instance <- function(df){
  # 1. Identify columns with the substring "Comment"
  comment_columns <- grep("Comment", names(df), value = TRUE)
  # 2. Extract substring before "Comment"
  prefix <- sub("(.*)Comment.*", "\\1", comment_columns)
  # Create a placeholder dataframe for merging
  merging_comment_df <- df[, comment_columns, drop = FALSE]
  # 3. Merge columns with the same prefix in merging_comment_df
  for (p in unique(prefix)) {
    cols_to_merge <- comment_columns[startsWith(comment_columns, p)]
    merging_comment_df[p] <- apply(df[cols_to_merge], 1, function(row) paste(row, collapse = ""))
  }
  # 4. Create a new dataframe new_df
  new_df <- df
  # 5. Match merging_comment_df back into new_df based on the last instance
  for (p in unique(prefix)) {
    last_instance_col <- tail(comment_columns[startsWith(comment_columns, p)], 1)
    new_df[last_instance_col] <- merging_comment_df[p]
  }
  # 6. Remove non-last instance of each unique substring in new_df
  for (p in unique(prefix)) {
    cols_to_remove <- setdiff(comment_columns[startsWith(comment_columns, p)], tail(comment_columns[startsWith(comment_columns, p)], 1))
    new_df <- new_df[, !names(new_df) %in% cols_to_remove, drop = FALSE]
  }
  # 7. retrurn the modified dataframe
  return(new_df)
}
```
### 2. function to Insert underscore between rowname and column name in flattened table to easier split and pivot later
```{r}
add_underscore_between_rowname_columnname <- function(df,example_prefix){
  #example_prefix is a regex string, and thus escape code(\\) are needed if () and other escape characters are included
  example_prefix = paste0("^",example_prefix," ") #^specifies start of the pattern, to avoid cases like *vaccinated* and un*vaccinated*
  #Change row-column separator of flattened table from default " " to "_" 
  suffixes <- colnames(df)[
  grepl(example_prefix, colnames(df))] %>% #examples "{example_prefix} {suffix}"
  sub(example_prefix,"",.) #remove prefix to get {suffix} only
  # Replace the space before the suffix in column names with underscores for each suffix for easier splitting later
  for (substring in suffixes) {
  df <- setNames(df, gsub(paste0(" ", substring), paste0("_", substring), colnames(df),fixed = TRUE))
  }
 return(df)
}
```
### 3. function serires to perform first pivots and filter 
pivot long, separate by _, then pivot wide again to make original table rownames (e.g. Cancer types, treatment, metastasis status) unique columns, like in data collection template
```{r}
first_pivot_long <- function(df_with_unique_delimiter_between_prefix_and_suffix,sep="_"){
  # first perform a long pivot, to get the 2 columns:
  #prefix (original table row names) and suffix (original table column names)
  long_data <- df_with_unique_delimiter_between_prefix_and_suffix %>% 
    gather(key = "key", value = "value", -`Covidence #`)
  #separate wider so suffixes can be pivoted
  long_data <- separate(long_data, key, into = c("prefix", "suffix"), sep = sep) 
  return (long_data)
}

first_pivot_wide <- function(long_data){
  # then pivot wide, to get 26 columns of recorded outcomes 
  wide_data <- long_data%>%
    pivot_wider(id_cols=c(`Covidence #`,prefix),names_from = suffix,values_from = value)
  return(wide_data)
}
first_pivoted_wide_filter <- function(wide_data){
  # filter away rows where no values are provided
  wide_data_filtered <- wide_data %>%
    # change all "" to NA, spare column 1 which will be Covidence #
    mutate_at(2:(ncol(wide_data)), ~ifelse(. == "", NA, .)) %>%
    # retain as long as any of the important values is not NA. 1-3 are Covidence #, prefix, Total Number of Patients, last one is Comment.1, and thus are spared in this filter
    filter(if_any(4:(ncol(wide_data)-1), ~(!is.na(.))))
  return(wide_data_filtered)
}
```
### 4. Function series to perform SECOND pivots and filter
pivot long, and then wide again
To pool different outcomes (mortality, ICU admission, severity etc.) onto a column so we can filter as such
```{r}
second_pivot_long_wide_filter <- function(df,
                              ID_columns=c("Covidence #","prefix",
                                           "Total Number of Patients","Comment.1"),
                              to_filter=TRUE){
  print(ID_columns)
  #convery spacer between colname and outcome to _ for easier separation
  colnames(df) <- gsub(pattern=" \\[",replacement="_\\[",colnames(df))
  #pivot long
  df_long2 <- df %>%
    gather(key="key",value="value",
           -ID_columns) %>%
    #separate by underscore
    separate(key,into=c("measurement","outcome"),sep="_")
  #then pivot wider
  df_wide2 <- df_long2 %>%
    pivot_wider(id_cols=c(ID_columns,"outcome"),
                names_from = measurement,values_from=value)
  #optionally filter
  if(to_filter==TRUE){
    #change all empty "" to NA
    # retain as long as any of the important values (MoA, MoA value, p-value, proportion) is not NA
    print(length(ID_columns))
    df_wide2_filtered <- df_wide2 %>% mutate_at(2:(ncol(df_wide2)),~ifelse(. =="",NA,.)) %>%
      filter(if_any((length(ID_columns)+2):ncol(df_wide2),~(!is.na(.))))
    
    return(df_wide2_filtered)
  }else{
    return(df_wide2)
  }
}
```
### 5. Check if formating is correct - same number of semicolons where necessary
```{r}
semi_colon_check <- function(df,
                             ID_columns){
  sparing_columns = (0:length(ID_columns)+1)
  df_semi_colons_count <- df %>%
    mutate_at(vars(-sparing_columns),~str_count(as.character(.),";")) %>%
  rowwise() %>% 
  mutate(same_number_of_semicolons=
           all(c_across(-sparing_columns))==first(c_across(-sparing_columns))
         )#find which rows have the same number of semicolons as thus can be properly delimiter splitted
  final_df = merge(df,df_semi_colons_count,by = c(ID_columns,"outcome"),
                   suffixes = c("","_number_of_semicolons"))
  return(final_df)
}
```
## Acting on Data Tables

### For Primary cancer sites
Covidence ID + each row is 26 items (1 total + 4 outcomes x6 measurements + 1 comment)
```{r}
primary_site_start_col <- which(colnames(df)=="Primary Cancer Site:") # column name before the comorbidity table
primary_site_end_col <- which(colnames(df)=="Metastasis status:") # column name before the comorbidity table

#1st clean, pivot and filter
df_1site <- df %>%
  select(all_of(c(1, #keep Covidence ID
                  (primary_site_start_col+1):(primary_site_end_col-1)
                  ))) %>%
  collapse_comment_to_last_instance() %>%
  add_underscore_between_rowname_columnname(example_prefix = "Thoracic \\(lung, mesothelioma\\)") %>%
  first_pivot_long() %>%
  first_pivot_wide() %>%
  first_pivoted_wide_filter()
  
df_1site
#2nd pivot, filter
df_1site_wide2_filtered<- second_pivot_long_wide_filter(df_1site,ID_columns = c(
  "Covidence #","prefix","Total Number of Patients","Comment.1"),
  to_filter = TRUE)
df_1site_wide2_filtered


## Total number of patients for Abe
df %>%
  select(all_of(c(1,2, #keep Covidence ID
                  (primary_site_start_col+1):(primary_site_end_col-1)
                  ))) %>%
  select(all_of(c(1,2,
                  contains("patients",ignore.case=TRUE)))) %>%
  write.csv("../output/abe_tables_graphs/CancerTypeTable_forAbe.csv")
```


```{r}
# sanity check - which papers do NOT have TOTAL cancers mortality
df$`Covidence #`[!df$`Covidence #` %in% df_1site_wide2_filtered[
  (df_1site_wide2_filtered$prefix == "Total (OF ALL CANCERS)") & 
    (df_1site_wide2_filtered$outcome == "[mortality]")
   ,]
  $`Covidence #`]
```
128789 focuses on multiple myeloma.
81620 describes a mixed outcome - outcomes are defined as death, or admission to hiospital. So there is no mortality data.
79599 no description of mortality.
75065 describes a composite outcome of composite endpoint of requirement for intubation, ICU admission and 30-day mortality.
73383 splits into solid and hematalogical, with no calculations of mixed/Total mortality.
72749,72735, haematological malignancies.
*62340 values stored at Other - as total patients have patients with ct value >32. Currently being fixed as our criteria is Ct below 35.*


### Third step: Adding the univariate/multivariate (measure of association) column by ; delimiters
#### at the moment not there yet due to unequal number of semicolons

```{r}
df_1site_wide2_filtered_semicolon_count <- semi_colon_check(
  df_1site_wide2_filtered,ID_columns = c("Covidence #","prefix",
                                         "Total Number of Patients","Comment.1")
  )
df_1site_wide2_filtered_semicolon_count

df_1site_wide2_filtered_semicolon_count %>% count(same_number_of_semicolons) # summarise count how many true, false, NA

write.csv(df_1site_wide2_filtered_semicolon_count,
          "../output/processed_tables/primary_cancer_sites_corrections.csv",row.names = FALSE)
```

## Initial separate and export in cases which can be succesfully separate_longer_delim
## 7. Actually perform semi-colon row splits, then widen columns by event/total proportions
IMPORTANT DO NOT DELETE
```{r}
last_semi_colon_split <- function(df,
                                  ID_columns){
  sparing_columns = length(ID_columns)+1 #+1 to prevent catching [outcome]
  final_df <- df %>%
    separate_longer_delim((sparing_columns+1):(sparing_columns+1+5),delim=";")
  #5 columns are measure of association, value,  95% CI, p-value, cancer: Event/Total, Non-cancer: Event/Total
  return (final_df)
}

widen_decimals_fractions <- function(df,cancer_proportion_col,control_proportion_col){
  df %>% mutate(
    "cancer(event/total)"=as.numeric(str_extract(df[[cancer_proportion_col]],"[[:digit:]]+\\.[[:digit:]]+")),
    "cancer(event count)"=as.integer(str_extract(df[[cancer_proportion_col]],"([[:digit:]]+(?=\\/))")),
    "cancer(total count)"=as.integer(str_extract(df[[cancer_proportion_col]],"((?<=\\/)[[:digit:]]+)"))
    ) %>%
    mutate(
    "non-cancer(event/total)"=as.numeric(str_extract(df[[control_proportion_col]],"[[:digit:]]+\\.[[:digit:]]+")),
    "non-cancer(event count)"=as.integer(str_extract(df[[control_proportion_col]],"([[:digit:]]+(?=\\/))")),
    "non-cancer(total count)"=as.integer(str_extract(df[[control_proportion_col]],"((?<=\\/)[[:digit:]]+)"))
    )
}

df_1site_wide2_filtered_semicolon_count_splitted <- last_semi_colon_split(
  df_1site_wide2_filtered_semicolon_count,
  ID_columns = c("Covidence #","prefix",
                 "Total Number of Patients","Comment.1")
  ) %>%
  widen_decimals_fractions(cancer_proportion_col = "cancer: Event/Total",
                         control_proportion_col = "Non-cancer: Event/Total")
df_1site_wide2_filtered_semicolon_count_splitted
write.csv(df_1site_wide2_filtered_semicolon_count_splitted,file = "../output/processed_tables/primary_cancer_sites_AnalysisReady.csv",row.names = FALSE)
```
## Metastasis status
```{r}
metastasis_start_col <- which(colnames(df)=="Metastasis status:") # column name before the mets table
metastasis_end_col <- which(colnames(df)=="Treatment type:") # column name before the mets table

df_metastasis <- df %>%
  select(all_of(c(1, #keep Covidence ID
                  (metastasis_start_col+1):(metastasis_end_col-1)))) %>%
  collapse_comment_to_last_instance() %>%
  add_underscore_between_rowname_columnname(example_prefix = "Unknown") %>%
  first_pivot_long() %>%
  first_pivot_wide() %>%
  first_pivoted_wide_filter()
df_metastasis

mets_ID_columns <- c("Covidence #","prefix","Total number of Patients","Comment.1")
df_metastasis <- second_pivot_long_wide_filter(df_metastasis,
                                               ID_columns = mets_ID_columns,
                                               to_filter = TRUE)
df_metastasis

df_metastasis <- semi_colon_check(
  df_metastasis,ID_columns = mets_ID_columns
  )
df_metastasis
df_metastasis %>% count(same_number_of_semicolons)
write.csv(df_metastasis,
          "../output/processed_tables/metastasis_status_corrections.csv",row.names = FALSE)

df_metastasis_final_split <- last_semi_colon_split(
  df_metastasis,ID_columns = mets_ID_columns
  ) %>%
  widen_decimals_fractions(cancer_proportion_col = "cancer: Event/Total",
                           control_proportion_col = "Non-cancer: Event/Total")
df_metastasis_final_split

write.csv(df_metastasis_final_split,
          "../output/processed_tables/metastasis_status_AnalysisReady.csv",row.names = FALSE)
```


### Treatment type - 27 columns as repeating unit instead
EXTRA COLUMN (1 total + *1 PERCENTAGE* +  4 outcomes x6 measurements + 1 comment) means everything repeats every 27 columns, rather than 26 columns like previous
```{r}
treatment_start_col <- which(colnames(df)=="Treatment type:") # column name before the treatment table
treatment_end_col <- which(colnames(df)=="COVID variants:") # column name before the treatment table
treatment_ID_cols <- c("Covidence #","prefix","Total number of patients","percentage","Comment")


df_treatment <- df %>%
  select(all_of(c(1, #keep Covidence ID
                  (treatment_start_col+1):(treatment_end_col-1)))) %>%
  collapse_comment_to_last_instance() %>%
  add_underscore_between_rowname_columnname(example_prefix="Chemotherapy") %>%
  first_pivot_long() %>%
  first_pivot_wide() %>%
  first_pivoted_wide_filter()

df_treatment

df_treatment<- second_pivot_long_wide_filter(df_treatment,
                                             ID_columns = treatment_ID_cols,
                                             to_filter = TRUE)
df_treatment

df_treatment <- semi_colon_check(df_treatment,ID_columns = treatment_ID_cols)
df_treatment
df_treatment %>% count(same_number_of_semicolons)
write.csv(df_treatment,"../output/processed_tables/treatment_type_corrections.csv",row.names=FALSE)

df_treatment_final_split <- last_semi_colon_split(
  df_treatment,ID_columns = treatment_ID_cols
  ) %>%
  widen_decimals_fractions(cancer_proportion_col = "cancer: Event/Total",
                           control_proportion_col = "Non-cancer: Event/Total")
df_treatment_final_split

write.csv(df_treatment_final_split,
          "../output/processed_tables/treatment_status_AnalysisReady.csv",row.names = FALSE)

## Total number of patients for Abe
df %>%
  select(all_of(c(1,2, #keep Covidence ID
                  (treatment_start_col+1):(treatment_end_col-1)
                  ))) %>%
  select(all_of(c(1,2,
                  matches('patients|comment|total number',ignore.case=TRUE)))) %>%
  write.csv("../output/abe_tables_graphs/CancerTreatmentTable_forAbe.csv")
```

### COVID variants - 27 columns as repeating unit instead
EXTRA COLUMN (1 total + *1 PERCENTAGE* +  4 outcomes x6 measurements + 1 comment) means everything repeats every 27 columns, rather than 26 columns like previous
```{r}
variant_start_col <- which(colnames(df)=="COVID variants:") # column name before the variant table
variant_end_col <- which(colnames(df)=="COVID vaccination status:") # column name before the variant table
variant_ID_columns = c("Covidence #","prefix","Total Number of patients","Percentage","Comment.1")

df_variant <- df %>%
  select(all_of(c(1, #keep Covidence ID
                  (variant_start_col+1):(variant_end_col-1))))
colnames(df_variant) <- colnames(df_variant)%>% gsub(
  pattern = "Measure of Association \\(OR/RR etc.\\) \\{severe\\]",
  replacement = "Measure of Association (OR/RR etc.) [severe]", 
  ) %>% gsub(
    pattern="\\.[0-9]+$",
    replacement="",
  ) # fix required for typo { instead of [
  
df_variant <- df_variant %>%
  collapse_comment_to_last_instance() %>% #first time to reassign and regain suffix for repeating Comment
  collapse_comment_to_last_instance() %>% #second time to truly collapse
  add_underscore_between_rowname_columnname(example_prefix = "Beta") %>%
  first_pivot_long() %>%
  first_pivot_wide() %>%
  first_pivoted_wide_filter()

df_variant

df_variant <- second_pivot_long_wide_filter(df_variant,ID_columns = variant_ID_columns,
                                            to_filter = TRUE)
df_variant

df_variant <- semi_colon_check(df_variant,ID_columns = variant_ID_columns)
df_variant
df_variant %>% count(same_number_of_semicolons)
write.csv(df_variant,"../output/processed_tables/variants_corrections.csv",row.names=FALSE)

df_variant_final_split <- last_semi_colon_split(
  df_variant,ID_columns = variant_ID_columns
  )%>%
  widen_decimals_fractions(cancer_proportion_col = "cancer: Event/Total",
                           control_proportion_col = "Non-cancer: Event/Total")
df_variant_final_split

write.csv(df_variant_final_split,"../output/processed_tables/variants_AnalysisReady.csv")
```

### vaccination status
```{r}
vaccination_start_col <- which(colnames(df)=="COVID vaccination status:") # column name before the vaccination table
vaccination_end_col <- which(colnames(df)=="Is individual patient data available?") # column name before the vaccination table
vaccination_IDcols <- c("Covidence #","prefix","Total number of patients","Percentage","Comment")

df_vaccination <- df %>%
  select(all_of(c(1, #keep Covidence ID
                  (vaccination_start_col+1):(vaccination_end_col-1)
                  ))) %>%
  collapse_comment_to_last_instance() %>%
  add_underscore_between_rowname_columnname(example_prefix = "vaccinated") %>%
  first_pivot_long()%>%
  first_pivot_wide() %>%
  first_pivoted_wide_filter()

df_vaccination

df_vaccination <- second_pivot_long_wide_filter(df_vaccination,
                                                ID_columns = vaccination_IDcols)
df_vaccination
df_vaccination <- semi_colon_check(df_vaccination,ID_columns = vaccination_IDcols)
df_vaccination
df_vaccination %>% count(same_number_of_semicolons)
write.csv(df_vaccination,"../output/processed_tables/vaccination_corrections.csv",row.names=FALSE)

df_vaccination_final_split <-
  last_semi_colon_split(df_vaccination,ID_columns=vaccination_IDcols
    ) %>%
  widen_decimals_fractions(cancer_proportion_col = "cancer: Event/Total",
                           control_proportion_col = "Non-cancer: Event/Total")
df_vaccination_final_split

write.csv(df_vaccination_final_split,"../output/processed_tables/vaccination_AnalysisReady.csv")
```

#Grouping effect sizes and preparation for meta-analysis and meta-regression

## step 1: rbinding primary cancer site, treatment type, and metastasis status. 

```{r}
# add in the percentage column, which exists in df_treatment and onwards but not in df_1site and df_metastasis
df_1site_final_to_stack <- cbind(df_1site_wide2_filtered_semicolon_count_splitted[1:3],percentage=NA,
                                 df_1site_wide2_filtered_semicolon_count_splitted[4:ncol(df_1site_wide2_filtered_semicolon_count_splitted)])
df_metastasis_final_to_stack <- cbind(df_metastasis_final_split[1:3],percentage=NA,
                                 df_metastasis_final_split[4:ncol(df_metastasis_final_split)])
# rename them so they share the same colnames
colnames(df_1site_final_to_stack) <- colnames(df_treatment_final_split)
colnames(df_metastasis_final_to_stack) <- colnames(df_treatment_final_split)
df_1site_metastasis_treatment <- bind_rows(df_1site_final_to_stack, df_metastasis_final_to_stack, df_treatment_final_split)

# Identify columns with the substring "_semicolons"
columns_to_remove <- grep("_semicolons", names(df_1site_metastasis_treatment))

# Remove columns with the substring "_semicolons"
df_1site_metastasis_treatment <- df_1site_metastasis_treatment%>%select(-all_of(columns_to_remove))%>%
  mutate(across(where(is.character), trimws))  # trim whitesapces if is.character
  
# if the event/total proportion can be calculated but is not present, add it.
df_1site_metastasis_treatment
df_1site_metastasis_treatment <- df_1site_metastasis_treatment %>% 
  mutate(`cancer(event/total)`=if_else(condition = is.na(df_1site_metastasis_treatment$`cancer(event/total)`),
                                       as.numeric(df_1site_metastasis_treatment$`cancer(event count)`)/as.numeric(df_1site_metastasis_treatment$`cancer(total count)`),
                                       as.numeric(df_1site_metastasis_treatment$`cancer(event/total)`)
                                       )
         ) %>%
  mutate(`non-cancer(event/total)`=if_else(condition = is.na(df_1site_metastasis_treatment$`non-cancer(event/total)`),
                                       as.numeric(df_1site_metastasis_treatment$`non-cancer(event count)`)/as.numeric(df_1site_metastasis_treatment$`non-cancer(total count)`),
                                       as.numeric(df_1site_metastasis_treatment$`non-cancer(event/total)`)
                                       )
         )
# correct instances where % are used instead of decimal
df_1site_metastasis_treatment %>%
  mutate(`cancer(event/total)`=if_else(condition = df_1site_metastasis_treatment$`cancer(event/total)` > 1.1, # correct instances where people put down % rather than 
                                       df_1site_metastasis_treatment$`cancer(event/total)`/100,
                                       df_1site_metastasis_treatment$`cancer(event/total)`
                                       )
         ) %>%
  mutate(`non-cancer(event/total)`=if_else(condition = df_1site_metastasis_treatment$`non-cancer(event/total)` > 1.1, # correct instances where people put down % rather than 
                                       df_1site_metastasis_treatment$`non-cancer(event/total)`/100,
                                       df_1site_metastasis_treatment$`non-cancer(event/total)`
                                       )
         )
# calculate cOR where possible (event/total)
cOR_required_condition <- is.na(df_1site_metastasis_treatment$value) ## find where OR/HR value is NA 

df_1site_metastasis_treatment <- df_1site_metastasis_treatment %>% #calculate OR
  mutate(value=ifelse(is.na(value),
                      `cancer(event/total)`/`non-cancer(event/total)`,
                      value)) %>%
  mutate(`Measure of Association` = ifelse((cOR_required_condition & # was original NA
                                                    !is.na(value) # but was able to yield a crude OR
                                                  ),
                                           "calculatedOR", #call its measure of association cOR
                                           `Measure of Association`) #otherwise keep as it was 
           )
# calculate standard error (se)
df_1site_metastasis_treatment <- df_1site_metastasis_treatment %>%
  mutate(se=sqrt(1/`cancer(event count)`+
                   1/`cancer(total count)`+
                   1/`non-cancer(event count)`+
                   1/`non-cancer(total count)`
                 )
         )
# add in if univariate or multivariate or if ambiguous
df_1site_metastasis_treatment <- df_1site_metastasis_treatment %>% mutate(
  model=case_when(
    str_detect(tolower(`Measure of Association`),"calculated|crude|(u|c|^)(or|rr|hr)") ~"univariate",
    str_detect(tolower(`Measure of Association`),"(m|a)(or|rr|hr)") ~"multivariate",
    TRUE~"manualclarificationrequired"
  )
)
df_1site_metastasis_treatment
```

```{r}
#adding in variants and vaccination of each study
df_vaccination_percentages <- df %>%
  select(all_of(c(1, #keep Covidence ID
                  (vaccination_start_col+1):(vaccination_end_col-1)
                  ))) %>%
  collapse_comment_to_last_instance() %>%
  select(`Covidence #`,`vaccinated Total number of patients`,`vaccinated Percentage`,`vaccinated Comment`,
         `unvaccinated Total number of patients`,`unvaccinated Percentage`,`unvaccinated Comment`)
df_vaccination_percentages
write.csv(df_vaccination_percentages,"../output/processed_tables/vaccination_status_percentages_manualchanges.csv",row.names = FALSE)
# manual correction
df_vaccination_percentages<- readxl::read_excel("../output/processed_tables/vaccination_status_percentages_manualchanges.xlsx")
df_vaccination_percentages
```

```{r}
df_variant <- df %>%
  select(all_of(c(1, #keep Covidence ID
                  (variant_start_col+1):(variant_end_col-1))))
colnames(df_variant) <- colnames(df_variant)%>% gsub(
  pattern = "Measure of Association \\(OR/RR etc.\\) \\{severe\\]",
  replacement = "Measure of Association (OR/RR etc.) [severe]", 
  ) %>% gsub(
    pattern="\\.[0-9]+$",
    replacement="",
  ) # fix required for typo { instead of [
df_variant <- df_variant %>%
  collapse_comment_to_last_instance() %>% #first time to reassign and regain suffix for repeating Comment
  collapse_comment_to_last_instance() #second time to truly collapse

desired_variant_columns <- grep("(Total Number of patients|Percentage)",names(df_variant),value=TRUE)
  
df_variant_selected <- df_variant[,c("Covidence #",desired_variant_columns)]
df_variant_selected
```

```{r}
#https://github.com/nextstrain/ncov-clades-schema?tab=readme-ov-file
#https://nextstrain.org/ncov/gisaid/global/6m
clade_names <- read.csv("../data/lineage/Nextstrain_clade_alias.csv",skip = 3)
#manual corrections
clade_names <- clade_names %>%
  mutate(variant=case_when(
    common_name==""~"Early-clades",
    startsWith(common_name,"XBB")~"Omicron_XBB",
    startsWith(common_name,"BA")~"Omicron_BA",
    common_name=="Omicron,B.1.1.628"~"Omicron_BA",
    common_name=="BQ.1"~"Omicron_BA",
    common_name=="CH.1.1"~"Omicron_BA",
    common_name=="EG.5.1"~"Omicron_XBB",
    common_name=="HK.3"~"Omicron_XBB",
    TRUE~common_name
  )) %>%
  arrange(nextstrain_clade)

desired_variant_order <- unique(clade_names$variant)
```


```{r}
#imputed variant information
assigned_lineages <- read.csv("../output/Timeline_and_Lineage_Imputation/studies_LAPIS_Outbreak_variant_percentage.csv",check.names = FALSE)
assigned_lineages <- assigned_lineages %>%
  group_by(`Study ID`,source) %>%
  slice(which.max(freq))%>%
  group_by(`Study ID`)%>%
  summarise(
    dateFrom=toString(unique(na.omit(dateFrom))),
    dateTo=toString(unique(na.omit(dateTo))),
    imputation_source=toString(unique(source)),
    imputed_variant=toString(unique(variant)),
    LAPIS_sumcount=sum(sumcount,na.rm = TRUE),
    Outbreak_count_total=sum(Outbreak_count_total,na.rm = TRUE)) %>%
  separate(`Study ID`,into=c("Covidence #","Study ID"),sep="_")
assigned_lineages
```

```{r}
merged_characteristics_regression <- merge(assigned_lineages,by.x ="Covidence #",
                                           df_variant_selected,by.y="Covidence #")
merged_characteristics_regression <- merge(merged_characteristics_regression,by.x="Covidence #",
                                           df_vaccination_percentages,by.y = "Covidence #",all = TRUE)

df_baseline_characteristics_selected <- df_baseline_characteristics

merged_characteristics_regression <- merge(merged_characteristics_regression,df_baseline_characteristics_selected,by="Covidence #")

merged_characteristics_regression
write.csv(merged_characteristics_regression,file = "../output/processed_tables/ForJAMES_CharacteristicsForRegression_ByStudy.csv")
```

```{r}
merged_characteristics_regression_withValues <- merge(merged_characteristics_regression,
      df_1site_metastasis_treatment,
      by="Covidence #")

merged_characteristics_regression_withValues
```


```{r}
df_comorbidity_percentages <- df_comorbidity[,c(1,
                  grep("Percentage",colnames(df_comorbidity)))]

merged_characteristics_regression_withValues_withComorb <- merge(
  merged_characteristics_regression_withValues,df_comorbidity_percentages,by="Covidence #")

merged_characteristics_regression_withValues_withComorb
write.csv(merged_characteristics_regression_withValues_withComorb,"../output/processed_tables/ForJAMES_CharacteristicsForRegression_ByStudy_withComorbs.csv")

merged_characteristics_regression_withValues_withComorb_new <- merged_characteristics_regression_withValues_withComorb %>% filter(`Covidence #` %in% c("70019","71402","71915"))
merged_characteristics_regression_withValues_withComorb_new
write.csv(merged_characteristics_regression_withValues_withComorb_new,"../output/processed_tables/70019_71402_71915/ForJAMES_CharacteristicsForRegression_ByStudy_withComorbs.csv")

```

```{r}
merged_characteristics_regression_withValues_withComorb %>% filter(`Study ID`=="Hosseini-Moghaddam 2023")
```
